---
title: "Exploratory Plots to Characterise the Dataset"
author: "Caitlin M"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: True
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "/Volumes/phd")

library(ggplot2)
library(tidyverse)
library(data.table)
library(VennDiagram)
library(UpSetR)
```

### This code is an exploration between the relationship between the variants found in the different populations as well as the overlap between various phenotype groups (as defined by the FANTOM consortium).
####This exploratory analysis will reveal underlying patterns in the data that may not be immediately evident and provide a better understanding of the data.


```{r files}
# Read in population files and give them column names
AFR <- fread("1KGP_hg38/AFR_variants_sorted.vcf", header = F)
AMR <- fread("1KGP_hg38/AMR_variants_sorted.vcf", header = F)
EAS <- fread("1KGP_hg38/EAS_variants_sorted.vcf", header = F)
EUR <- fread("1KGP_hg38/EUR_variants_sorted.vcf", header = F)
SAS <- fread("1KGP_hg38/SAS_variants_sorted.vcf", header = F)

# Assign consistent column names
colnames(AFR) <- colnames(AMR) <- colnames(EAS) <- colnames(EUR) <- colnames(SAS) <- 
  c("CHR", "start_pos", "end_pos", "strand", "REF", "ALT", "AF") 

# Create unique SNP IDs (e.g., "chr1:12345")
variant_lists <- list(
  AFR = paste0("chr", AFR$CHR, ":", AFR$start_pos),
  AMR = paste0("chr", AMR$CHR, ":", AMR$start_pos),
  EAS = paste0("chr", EAS$CHR, ":", EAS$start_pos),
  EUR = paste0("chr", EUR$CHR, ":", EUR$start_pos),
  SAS = paste0("chr", SAS$CHR, ":", SAS$start_pos)
)
```
##Number of Variants per Population
#### These plots look at the the number of variants per population. There is one plot that is the total number of variants in the population and the second plot is the number of unique variants - variants that only appear in one population - for each superpopulation

```{r pop}

# read in files for the total number of variants for each superpopulation
pop <- read.table("1KGP_hg38/pop_count_hg38.txt", sep = "\t")
colnames(pop) <- c("Superpopulation", "Total_Lines")
bootstrapped <- read.table("prelim_exploration/bootstrap_results.txt", sep = "\t", header = T) # bootstrapping results for the total variant count
pop_boot <- full_join(pop, bootstrapped, by = "Superpopulation")
# adjusting the lines so they are not in scientific notation
pop_boot$Adjusted_lines <- pop_boot$Total_Lines / 1e7
pop_boot$Adjusted_uCI <- pop_boot$Upper_CI / 1e7
pop_boot$Adjusted_lCI <- pop_boot$Lower_CI / 1e7


# read in files for the  number of variants unique to each superpopulation
unique_pop <- read.table("1KGP_hg38/unique_pop_count_hg38.txt", sep = "\t")
colnames(unique_pop) <- c("Superpopulation", "Lines")
unique_bootstrap <- read.table("prelim_exploration/unique_bootstrap_results.txt", sep = "\t", header = T)
uniq_pop_boot <- full_join(unique_pop, unique_bootstrap, by = "Superpopulation")
# adjusting the lines so they are not in scientific notation
uniq_pop_boot$Adjusted_lines <- uniq_pop_boot$Lines / 1e6
uniq_pop_boot$Adjusted_uCI <- uniq_pop_boot$Upper_CI / 1e6
uniq_pop_boot$Adjusted_lCI <- uniq_pop_boot$Lower_CI / 1e6

# plotting total variants per population
pop_plot <- ggplot(pop_boot, aes(Superpopulation, Adjusted_lines, fill = Superpopulation)) + geom_bar(stat="identity") +
  geom_errorbar(aes(x = Superpopulation, ymin = Adjusted_lCI, ymax = Adjusted_uCI)) +
  ylab(bquote("Number of Variants "(x10^7))) +
  xlab("Superpopulation") +
  theme_classic() +
  theme(legend.position="none")
pop_plot
# Save high-resolution image
#ggsave("population_variant_count.png", pop_plot, dpi = 300, width = 6, height = 4)

# plotting the number of unique variant per population
unique_pop_plot <- ggplot(uniq_pop_boot, aes(Superpopulation, Adjusted_lines, fill = Superpopulation)) + geom_bar(stat="identity") +
  geom_errorbar(aes(x = Superpopulation, ymin = Adjusted_lCI, ymax = Adjusted_uCI)) +
  ylab(bquote("Number of Unique Variants "(x10^6))) +
  xlab("Superpopulation") +
  theme_classic() +
  theme(legend.position="none")
unique_pop_plot
#ggsave("unique_population_variant_count.png", unique_pop_plot, dpi = 300, width = 6, height = 4)


```

## Variant per chromosome count
#### There are two plots that are looking at the number of variants per chromosome for each of the superpopulations. The first plot looks at the number of variants per chromosome, whereas the second plot looks at the the number of variants in the chromosome adjusted by the size of the chromosome.

```{r chrom_count}
# read in file with the number of positions in each chromosome
hg38 <-  read.table("hg38.genome", header = T)

# Function to process and normalise counts per population
get_normalised_counts <- function(df, name) {
  df %>%
    filter(CHR %in% hg38$CHR) %>%
    count(CHR, name = "Variant_Count") %>%
    left_join(hg38, by = "CHR") %>%
    mutate(Population = name,
           Variants_per_bp = Variant_Count / size)
}

# Apply to each population
afr_norm <- get_normalised_counts(AFR, "AFR")
amr_norm <- get_normalised_counts(AMR, "AMR")
eas_norm <- get_normalised_counts(EAS, "EAS")
eur_norm <- get_normalised_counts(EUR, "EUR")
sas_norm <- get_normalised_counts(SAS, "SAS")

# Combine all
all_norm <- bind_rows(afr_norm, amr_norm, eas_norm, eur_norm, sas_norm)

# Clean up chromosome factor levels
all_norm$CHR <- factor(all_norm$CHR, levels = c(1:22, "X"))

# Histogram of the number of variants per population 
ggplot(all_norm, aes(x = CHR, y = Variant_Count, fill = Population)) +
  geom_bar(stat = "identity") +
  facet_wrap(~Population, scales = "free_y") +
  ylab("Number of Variants") +
  xlab("Chromosome") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Histogram of the normalised counts per chromosome for each population
ggplot(all_norm, aes(x = CHR, y = Variants_per_bp, fill = Population)) +
  geom_bar(stat = "identity") +
  facet_wrap(~Population, scales = "free_y") +
  ylab("Normalized Variant Density (Variants per base pair)") +
  xlab("Chromosome") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r presence-matrix,echo=FALSE}
## Create Presence Matrix and Summary Table

# All unique SNPs across all populations
all_snps <- unique(unlist(variant_lists))

# Build binary matrix (presence/absence)
presence_matrix <- sapply(variant_lists, function(snps) all_snps %in% snps)
rownames(presence_matrix) <- all_snps

# Count in how many populations each SNP appears
population_count <- rowSums(presence_matrix)

# Create summary table
summary_table <- as.data.frame(table(population_count))
colnames(summary_table) <- c("Num_Populations", "Num_Variants")

# Save summary table
#write.table(summary_table, "snp_overlap_summary.txt", sep = "\t", row.names = FALSE, quote = FALSE)
```

## UpSet Plot for All Populations
#### This plot shows the relationship between the superpopulations by counting the number of variants that appear in the different ralationship and measuring the overlap between them. From this we can see which populations have the highest degree of relatedness

```{r upset-plot, fig.width=14, fig.height=10}
# Convert binary matrix to data frame for UpSetR
# Ensure binary (0/1) data frame
presence_df <- as.data.frame(presence_matrix)
presence_df <- presence_df %>%
  mutate(across(everything(), ~ as.integer(.)))

set_sizes <- colSums(presence_df)

# Sort population names by size (smallest to largest)
sorted_sets <- names(sort(set_sizes, decreasing = T))

set_colors <- c(
  AFR = "#F8766D",
  AMR = "#A3A500",
  EAS = "#00BF7D",
  EUR = "#00B0F6",
  SAS = "#E76BF3"
)

UpSetR::upset(
  presence_df,
  nsets = length(sorted_sets),
  sets = sorted_sets,
  order.by = "freq",
  sets.bar.color = set_colors[sorted_sets],
  main.bar.color = "#333333",          # Keep main bars neutral
  matrix.color = "black",              # Matrix dots remain black
  matrix.dot.alpha = 0.9,
  set_size.show = F,
  text.scale = c(1.7, 1.5, 1.2, 0, 1.3, 1.3),
  number.angles = 30,
  keep.order = TRUE
)

```

## Venn Diagrams of the Overlap with the European superpopulation
#### These venn diagrams look at the overlap between the European superpopulation and the other 4. 
#### We are looking at the European overlaps due to the extreme European bias in genome-wide association studies.

```{r venn-diagram, fig.width=6, fig.height=6}
library(grid)
# Create a function that generates a venn diagram between two superpopulations and asks for the colour to plot them with
plot_venn <- function(pop1, pop2, col1, col2){
  venn_obj <- venn.diagram(
    x = variant_lists[c(pop1, pop2)],
    category.names = c(pop1, pop2),  # pop1 appears on the left
    filename = NULL,
    output = TRUE,
    imagetype = "png",
    col = "transparent",
    fill = c(col1, col2),
    alpha = 0.5,
    cex = 1.2,
    cat.cex = 1.2,
    print.mode = c("raw", "percent"),
    percent = TRUE
  )
  
  grid.newpage()
  grid.draw(venn_obj)
}

```
###Venn Diagram of EUR and AFR
```{r ven_eur_afr, fig.width=5, fig.height=5}
plot_venn("EUR", "AFR", "#00B0F6", "#F8766D")
```
###Venn Diagram of EUR and AMR
```{r ven_eur_amr, fig.width=5, fig.height=5}
grid.newpage()
grid.draw(plot_venn("EUR", "AMR", "#00B0F6", "#A3A500"))
```
###Venn Diagram of EUR and EAS
```{r ven_eur_eas, fig.width=5, fig.height=5}
grid.newpage()
grid.draw(plot_venn("EUR", "EAS", "#00B0F6", "#00BF7D"))
```
###Venn Diagram of EUR and SAS
```{r ven_eur_sas, fig.width=5, fig.height=5}
grid.newpage()
grid.draw(plot_venn("EUR", "SAS", "#00B0F6", "#E76BF3"))
```
### Bar Plot of the number of populations that a variant is shared between
```{r snp-sharing, fig.width=6, fig.height=4}
ggplot(summary_table, aes(x = as.factor(Num_Populations), y = Num_Variants)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  xlab("Number of Populations Sharing Variant") +
  ylab("Number of Variantss") +
  theme_classic()
```

## Allele Frequency
###Histogram of the allele frequencies across different populations

```{r AF, echo=FALSE}
pops <- c('AFR', 'AMR', 'EAS', 'EUR', 'SAS')
af_list <- list()
af1 <- data.frame(Population = character(), AF_1_count = integer(), stringsAsFactors = FALSE)

for (pop in pops){
  af_data <- get(pop) %>% count(AF) %>% as.data.frame()
  assign(paste0("af_", pop), af_data)
  
  #af_list[[pop]] <- af_data
  af1_count <- af_data[af_data$AF == 1.00, "n"]
  af1 <- rbind(af1, data.frame(Population = pop, AF_1_count = af1_count))
}


afr_plot <- ggplot(af_AFR, aes(AF, n)) + geom_bar(stat = "identity", fill = "#F8766D") +
  xlab("Allele Frequency") +
  theme_classic()
afr_plot

amr_plot <- ggplot(af_AMR, aes(AF, n)) + geom_bar(stat = "identity", fill = "#A3A500") +
  xlab("Allele Frequency") +
  theme_classic()
amr_plot

eas_plot <- ggplot(af_EAS, aes(AF, n)) + geom_bar(stat = "identity", fill = "#00BF7D") +
  xlab("Allele Frequency") +
  theme_classic()
eas_plot

eur_plot <- ggplot(af_EUR, aes(AF, n)) + geom_bar(stat = "identity", fill = "#00B0F6") +
  xlab("Allele Frequency") +
  theme_classic()
eur_plot

sas_plot <- ggplot(af_SAS, aes(AF, n)) + geom_bar(stat = "identity", fill = "#E76BF3") +
  xlab("Allele Frequency") +
  theme_classic()
sas_plot
```

### Violin plots of the different allele frequencies within each chromosome

```{r af_chr}
library(ggridges)

ggplot(EUR, aes(y = CHR, x = AF, fill = CHR)) +
  geom_density_ridges(scale = 2, alpha = 0.8, color = "white") +
  labs(title = "Allele Frequency Distribution by Chromosome (EUR)",
       x = "Allele Frequency", y = "Chromosome") +
  theme_ridges() +
  theme(legend.position = "none")

# AFR 
ggplot(AFR, aes(x = CHR, y = AF)) +
  geom_violin(trim = FALSE, alpha = 0.7, fill = "#F8766D") +
  labs(title = "Allele Frequency Distribution per Chromosome (AFR)",
       x = "Chromosome", y = "Allele Frequency") +
  theme_minimal() +
  theme(legend.position = "none")

# AMR 
ggplot(AMR, aes(x = CHR, y = AF)) +
  geom_violin(trim = FALSE, alpha = 0.7, fill = "#A3A500") +
  labs(title = "Allele Frequency Distribution per Chromosome (AMR)",
       x = "Chromosome", y = "Allele Frequency") +
  theme_minimal() +
  theme(legend.position = "none")

# EAS 
ggplot(EAS, aes(x = CHR, y = AF)) +
  geom_violin(trim = FALSE, alpha = 0.7, fill = "#00BF7D") +
  labs(title = "Allele Frequency Distribution per Chromosome (EAS)",
       x = "Chromosome", y = "Allele Frequency") +
  theme_minimal() +
  theme(legend.position = "none")

ggplot(EUR, aes(x = CHR, y = AF)) +
  geom_violin(trim = FALSE, alpha = 0.7, fill = "#00B0F6") +
  labs(title = "Allele Frequency Distribution per Chromosome (EUR)",
       x = "Chromosome", y = "Allele Frequency") +
  theme_minimal() +
  theme(legend.position = "none")

# SAS 
ggplot(SAS, aes(x = CHR, y = AF)) +
  geom_violin(trim = FALSE, alpha = 0.7, fill = "#E76BF3") +
  labs(title = "Allele Frequency Distribution per Chromosome (SAS)",
       x = "Chromosome", y = "Allele Frequency") +
  theme_minimal() +
  theme(legend.position = "none")

```
## Fixed Alleles

```{r fixed_alleles}
# a for loop that gets the distribution of the allele frequencies for each population
pops <- c('AFR', 'AMR', 'EAS', 'EUR', 'SAS')
af_list <- list()
af1 <- data.frame(Population = character(), AF_1_count = integer(), stringsAsFactors = FALSE)

for (pop in pops){
  af_data <- get(pop) %>% count(AF) %>% as.data.frame()
  assign(paste0("af_", pop), af_data)
  
  #af_list[[pop]] <- af_data
  af1_count <- af_data[af_data$AF == 1.00, "n"]
  af1 <- rbind(af1, data.frame(Population = pop, AF_1_count = af1_count))
}

ggplot(af1, aes(Population, AF_1_count,fill = Population)) + geom_bar(stat = "identity") + 
  theme_classic() +
  theme(legend.position = "none") +
  xlab("Superpopulation") +
  ylab("Number of Variants")
  
  
```

## Population Phenotypes
#### These plots are generated from the bedtools summary of the overlap between superpopulation variants and phenotype groups. The second plot is adjusted for the number of variants in the superpopulation to account for the differences in the number of variants in different populations

```{r phenotype_dist}
# reading in the bedtools summary file and manipulating the rows so they can be plotted
lines <- read_lines("gwas_1000_genomes/bedtools_summary.txt")[-1] # reads in the lines of the file, excluding the first line
bedtools <- tibble(raw = lines) %>%
  mutate(raw = str_remove(raw, " lines$")) %>%  # Remove " lines" from the end
  separate(raw, into = c("Superpopulation", "Phenotype"), sep = "_vs_", extra = "merge") %>%  # Split at "_vs_"
  separate(Phenotype, into = c("Phenotype", "Line_Count"), sep = ": ") %>%  # Split at ": "
  mutate(Phenotype = str_remove(Phenotype, "_gwas$"),  # Remove "_gwas"
         Line_Count = as.integer(Line_Count))  # Convert count to integer

bed_boot <- read.table("prelim_exploration/bootstrap_superpop_phenotype_results.txt", sep = "\t", header = T)
bedtools_boot <- full_join(bedtools, bed_boot, by = c("Superpopulation", "Phenotype"))
bedtools_boot <- subset(bedtools_boot, select = -c(Line_Count))
# creating a new dataframe that merges the dataframes with the number of variants and the phenotypes so the proportion can be calculated
proportion_df <- full_join(pop, bedtools_boot, by ="Superpopulation")
proportion_df <- proportion_df[order(proportion_df$Superpopulation),]
proportion_df$Proportion <- proportion_df$Count / proportion_df$Total_Lines *100
proportion_df$Adjusted_uCI <- proportion_df$Upper_CI / proportion_df$Total_Lines *100
proportion_df$Adjusted_lCI <- proportion_df$Lower_CI / proportion_df$Total_Lines *100

phenotype_order <-  c('Cancer', 'Neurological', 'Immunological') # changing the order of the phenotypes so it is more aesthetic
# plotting the raw numbers of the overlap between the variants and the phenotype groups
bedtools_plot <- ggplot(bedtools_boot, aes(factor(Phenotype, level = phenotype_order), Count, fill = Superpopulation)) + 
  geom_bar(stat = "identity",position=position_dodge(), color="white",) + 
  geom_errorbar(aes(ymin = Lower_CI, ymax = Upper_CI), 
                position = position_dodge(width = 0.9), 
                width = 0.25, 
                color = "black") +
  xlab("Phenotype") +
  ylab("Number of Variants") +
  theme_classic()
bedtools_plot
#ggsave("phenotype_variant_count.png", bedtools_plot, dpi = 300, width = 7, height = 5)

# plotting the proportion of the phenotype overlap compared to the total number of variants for the population
prop_plot <- ggplot(proportion_df, aes(factor(Phenotype, level = phenotype_order), Proportion, fill = Superpopulation)) + 
  geom_bar(stat = "identity",position=position_dodge(), color="white",) + 
  geom_errorbar(aes(ymin = Adjusted_lCI, ymax = Adjusted_uCI), 
                position = position_dodge(width = 0.9), 
                width = 0.25, 
                color = "black") +
  xlab("Phenotype") +
  ylab("Propotion of the total variants (%)") +
  theme_classic()
prop_plot
#ggsave("phenotype_variant_proportion.png", prop_plot, dpi = 300, width = 7, height = 5)



```




```{r ancestry}

ancestry <- read.table("NHGRI_EBI_GWAS/gwas_catalog-ancestry_r2025-02-18.tsv", header = T, sep = "\t")


```
